---
title: "CharAnalysis: Fire-history reconstruction from lake-sediment charcoal"
author: "Philip Higuera"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{CharAnalysis: Fire-history reconstruction from lake-sediment charcoal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  warning   = FALSE,
  message   = FALSE
)
```

## Overview

**CharAnalysis** reconstructs local fire histories from lake-sediment charcoal records. Charcoal preserved in lake sediments is a direct proxy for past fire activity: individual fire events near a lake deposit pulses of charcoal that appear as peaks above a slowly varying background signal. CharAnalysis formalises this peak-detection logic as a reproducible, quantitative workflow.

This R package is a direct translation of CharAnalysis v2.0 (MATLAB), validated against reference outputs on the Code Lake benchmark dataset. Analytical methods are described in Higuera et al. (2009).

The full workflow proceeds in five steps:

1. **Pretreatment** — Resample the raw charcoal series to equal time steps and compute charcoal accumulation rate (CHAR, pieces cm^-2^ yr^-1^).
2. **Smoothing** — Estimate low-frequency trends (C~background~) using lowess or a moving-window statistic.
3. **Peak decomposition** — Compute the high-frequency residual C~peak~ = C~interp~ − C~background~ (or ratio C~interp~ / C~background~).
4. **Thresholding** — Define a noise threshold and flag C~peak~ values that exceed it as candidate fire events.  Thresholds can be global (one distribution fitted to the full record) or local (sliding-window distribution).
5. **Screening and metrics** — Apply a minimum-count significance test, then compute fire-return intervals (FRIs), Weibull statistics, smoothed fire frequency, signal-to-noise index (SNI), and goodness-of-fit.

---

## Installation

```{r install, eval = FALSE}
# Install from GitHub (requires devtools)
devtools::install_github("phiguera/CharAnalysis",
                         subdir = "CharAnalysis_2_0_R")

# Suggested packages for figures
install.packages(c("ggplot2", "patchwork", "ggtext"))
```

---

## Worked example: Code Lake

Code Lake (CO; Colorado, USA) is the primary validation dataset for CharAnalysis. The record spans approximately 7,300 calibrated years BP and is analysed here using a local Gaussian mixture model (GMM) threshold — the recommended default configuration.

### Input files

CharAnalysis reads two CSV files:

- **`CO_charData.csv`** — the charcoal data table (depth, age, volume, count).
- **`CO_charParams.csv`** — the analysis parameter file.

Both files are included in the package's `inst/extdata/` directory and can be located with `system.file()`:

```{r paths, eval = FALSE}
params_file <- system.file("extdata", "CO_charParams.csv",
                           package = "CharAnalysis")
```

For this vignette we reference the files directly from the repository:

```{r set-paths, eval = FALSE}
params_file <- "path/to/CO_charParams.csv"   # adjust to your local path
```

### Running the full pipeline

A single call to `CharAnalysis()` runs all five analytical steps and returns a named list of results.

```{r run, eval = FALSE}
library(CharAnalysis)

out <- CharAnalysis(params_file)
```

The progress messages show each step completing in sequence:

```
(1) Reading input file...
      ...done.
(1b) Validating input parameters...
(2) Pretreating charcoal data...
      ...done.
(3) Smoothing resampled CHAR to estimate low-frequency trends
    and calculating peak CHAR...
      ...done.
(4) Defining possible thresholds for peak identification...
      ...done.
(5) Identifying peaks and applying minimum-count screening...
      ...done.
(6) Post-processing: fire-return intervals, Weibull statistics...
      ...done.
(7) Pipeline complete.  Call CharWriteResults(...) to save output CSV.
```

### Exploring the output

`CharAnalysis()` returns a named list. The most commonly used elements are:

```{r output-structure, eval = FALSE}
names(out)
#> [1] "charcoal"      "pretreatment"  "smoothing"     "peak_analysis"
#> [5] "results"       "site"          "gap_in"        "char_thresh"
#> [9] "post"          "char_results"
```

#### The `charcoal` object

`out$charcoal` holds all time-series outputs, both raw and processed:

```{r charcoal, eval = FALSE}
# Inspect the first few rows of key time series
head(data.frame(
  age_BP    = out$charcoal$ybpI,          # interpolated age (yr BP)
  CHAR      = out$charcoal$accI,          # C_interpolated  (pieces cm-2 yr-1)
  C_bkg     = out$charcoal$accIS,         # C_background
  C_peak    = out$charcoal$peak,          # C_peak (residuals)
  peaks     = out$charcoal$charPeaks[, 4] # final-threshold peak flags (0/1)
))
```

#### The `char_thresh` object

`out$char_thresh` holds threshold values, SNI, and goodness-of-fit results:

```{r thresh, eval = FALSE}
# Threshold at the final percentile (column 4 = threshValues[4])
range(out$char_thresh$pos[, 4], na.rm = TRUE)

# Signal-to-noise index (SNI): values > 3 indicate a strong signal
summary(out$char_thresh$SNI)
```

#### Post-processing metrics

`out$post` holds fire-history summary metrics:

```{r post, eval = FALSE}
# Fire-return intervals (FRIs) and mean FRI
cat("Number of FRIs:", length(out$post$FRI), "\n")
cat("Mean FRI:", round(mean(out$post$FRI), 1), "yr\n")

# Per-zone Weibull statistics (zone 1)
fri_z1 <- out$post$FRI_params_zone[1, ]
cat(sprintf(
  "Zone 1 — nFRI: %d  mFRI: %.1f yr  WBLb: %.1f  WBLc: %.2f\n",
  fri_z1[1], fri_z1[2], fri_z1[5], fri_z1[8]
))
```

#### The `char_results` matrix

`out$char_results` is the complete N × 33 output matrix, with columns matching the MATLAB `charResults` CSV exactly:

```{r char-results, eval = FALSE}
dim(out$char_results)
#> [1] 500  33

# Total number of fire events identified
sum(out$charcoal$charPeaks[, 4], na.rm = TRUE)
#> [1] 39
```

---

### Output figures

CharAnalysis provides five publication-quality ggplot2 figures. All are produced by `char_plot_all()`, or individually by their dedicated functions.

#### Figure 3 — C~interp~, C~background~, and C~peak~

```{r fig3, eval = FALSE}
char_plot_char(out)
```

The top panel shows resampled CHAR (black bars) with the smoothed C~background~ trend (grey line). The bottom panel shows C~peak~ with the positive and negative threshold lines (red), identified fire events (+ symbols), and peaks that failed the minimum-count screen (grey dots).

#### Figure 5 — Cumulative peaks through time

```{r fig5, eval = FALSE}
char_plot_cumulative(out)
```

The slope of the cumulative curve at any point reflects the instantaneous fire frequency. Changes in slope indicate periods of higher or lower fire activity.

#### Figure 6 — FRI distributions by zone

```{r fig6, eval = FALSE}
char_plot_fri_dist(out)
```

Each panel shows a histogram of fire-return intervals within one stratigraphic zone (20-yr bins, normalised to proportions). A two-parameter Weibull probability density function is overlaid where the Kolmogorov-Smirnov goodness-of-fit test passes (p > 0.10 for n < 30; p > 0.05 for n ≥ 30). Weibull scale (b) and shape (c) parameters, mean FRI, and sample size are annotated.

#### Figure 7 — Continuous fire history

```{r fig7, eval = FALSE}
char_plot_fire_history(out)
```

Three stacked panels show (from top to bottom): peak magnitude (integrated C~peak~ above threshold per fire event, pieces cm^-2^ peak^-1^); fire-return intervals through time as filled squares with the smoothed mean FRI curve (black line) and bootstrapped 95% CI ribbon (grey); and lowess-smoothed fire frequency (fires per 1000 yr).

#### Figure 8 — Between-zone CHAR comparisons

```{r fig8, eval = FALSE}
char_plot_zones(out)
```

The left panel shows empirical cumulative distribution functions of raw CHAR within each zone, with pairwise two-sample Kolmogorov-Smirnov test p-values annotated. The right panel shows box plots (10th, 25th, 50th, 75th, 90th percentiles) of raw CHAR by zone.

#### Saving all figures to PDF

```{r save-figs, eval = FALSE}
char_plot_all(out, save = TRUE, out_dir = "Results")
# Saves:
#   Results/CO_03_CHAR_analysis.pdf
#   Results/CO_05_cumulative_peaks.pdf
#   Results/CO_06_FRI_distributions.pdf
#   Results/CO_07_continuous_fire_hx.pdf
#   Results/CO_08_zone_comparisons.pdf
```

---

### Writing results to CSV

`CharWriteResults()` writes the 33-column output matrix to a CSV file whose column names and format match the MATLAB `charResults` output exactly:

```{r write, eval = FALSE}
CharWriteResults(out$char_results,
                 site    = out$site,
                 out_dir = "Results")
# Writes: Results/CO_charResults.csv
```

The output CSV contains one row per interpolated time step and 33 columns covering all analytical outputs from C~interp~ through to per-zone Weibull confidence intervals. Column headers match the MATLAB reference file exactly to facilitate direct numerical comparison.

---

## Key parameters

The parameter file (`*_charParams.csv`) controls all aspects of the analysis. The most commonly adjusted parameters are:

| Parameter | Default | Description |
|-----------|---------|-------------|
| `yrInterp` | 15 | Resampling resolution (yr). Set to 0 for automatic (median raw resolution). |
| `yr` | 500 | Smoothing window width (yr) for C~background~ estimation. |
| `threshType` | 2 | Threshold type: 1 = global, 2 = local (sliding window). |
| `threshMethod` | 3 | Noise distribution: 2 = Gaussian, 3 = Gaussian mixture model. |
| `threshValues` | 0.95, 0.99, 0.999, 0.99 | Percentile thresholds; the final value defines the working threshold. |
| `minCountP` | 0.05 | Alpha level for the minimum-count significance screen. |
| `peakFrequ` | 1000 | Window width (yr) for smoothed fire frequency and FRI statistics. |
| `zoneDiv` | record bounds | Zone boundaries (yr BP) for stratified FRI and Weibull analysis. |

---

## Comparison with MATLAB v2.0

This R package is a direct translation of CharAnalysis v2.0 (MATLAB). Outputs are quantitatively equivalent on all validated reference datasets, with the following documented differences:

**Gaussian mixture model (GMM) thresholds** — The R package uses a direct port of the MATLAB `GaussianMixture.m` EM algorithm (`GaussianMixture.R`). Floating-point arithmetic accumulates differently across languages during EM iterations, causing the two implementations to converge to slightly different local solutions in some windows. This results in an approximately 20% difference in peak counts on the Code Lake dataset (39 R peaks vs. 48 MATLAB peaks). Threshold values themselves agree well (maximum difference < 0.094 pieces cm^-2^ yr^-1^).

**Weibull confidence intervals** — Bootstrap CIs are stochastic and will differ between R and MATLAB runs due to differences in random-number generation. Weibull point estimates (WBLb, WBLc) agree within < 5% on validated datasets.

**Zone 2 FRI statistics (Code Lake)** — Because R identifies fewer peaks in the older part of the Code Lake record, zone 2 has insufficient FRIs for Weibull fitting. This is a downstream consequence of the GMM floating-point divergence, not an implementation error.

---

## Citation

If you use CharAnalysis in published research, please cite Higuera et al. (2009), the first study to apply the core analytical tools implemented in the program. If you used Version 2.0 specifically, please also cite the software:

> Higuera, P.E., L.B. Brubaker, P.M. Anderson, F.S. Hu, and T.A. Brown. 2009. Vegetation mediated the impacts of postglacial climate change on fire regimes in the south-central Brooks Range, Alaska. *Ecological Monographs* 79:201–219. https://doi.org/10.1890/07-2019.1

> Higuera, P.E. 2026. *CharAnalysis*: Diagnostic and analytical tools for peak analysis in sediment-charcoal records (Version 2.0). Zenodo. https://doi.org/10.5281/zenodo.19304064

---

## References

Higuera, P.E., L.B. Brubaker, P.M. Anderson, F.S. Hu, and T.A. Brown. 2009. Vegetation mediated the impacts of postglacial climate change on fire regimes in the south-central Brooks Range, Alaska. *Ecological Monographs* 79:201–219. https://doi.org/10.1890/07-2019.1
