---
title: "CharAnalysis: Diagnostic and analytical tools for peak detection and fire-history interpretations using high-resolution sediment-charcoal records"
author: "Philip Higuera"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{CharAnalysis: Diagnostic and analytical tools for peak detection and fire-history interpretations using high-resolution sediment-charcoal records}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  warning   = FALSE,
  message   = FALSE
)

# Suggests-package gate: figure chunks live-render only when ggplot2,
# patchwork, and ggtext are all installed. If any is missing, those
# chunks skip cleanly and the rest of the vignette still builds.
PLOT_OK <- requireNamespace("ggplot2",   quietly = TRUE) &&
           requireNamespace("patchwork", quietly = TRUE) &&
           requireNamespace("ggtext",    quietly = TRUE)
```

## Overview

*CharAnalysis* is a tool to help reconstruct local fire history from lake-sediment charcoal records. Charcoal preserved in lake sediments is a direct proxy for past fire activity, and peaks in charcoal deposition above a slowly varying background level have been shown to record individual fire events near a lake. *CharAnalysis* formalizes this peak-detection logic as a reproducible, quantitative workflow.

This R package is a direct translation of *CharAnalysis* v2.0 (MATLAB), validated against reference outputs on four benchmark datasets spanning a range of record lengths, sampling resolutions, ecosystems, and analysis configurations. Analytical methods are described in Higuera et al. (2009).

The full workflow proceeds in five steps:

1.  **Pretreatment** — Resample the raw charcoal series to equal time steps and compute charcoal accumulation rate (CHAR, pieces cm^-2^ yr^-1^).
2.  **Smoothing** — Estimate low-frequency trends (C~background~) using lowess or a moving-window statistic.
3.  **Peak decomposition** — Compute the high-frequency residual C~peak~ = C~interp~ − C~background~ (or ratio C~interp~ / C~background~).
4.  **Thresholding** — Define a noise threshold and flag C~peak~ values that exceed it as candidate fire events. Thresholds can be global (one distribution fitted to the full record) or local (sliding-window distribution).
5.  **Screening and metrics** — Apply a minimum-count significance test, then compute fire-return intervals (FRIs), Weibull statistics, smoothed fire frequency, signal-to-noise index (SNI), and goodness-of-fit.

------------------------------------------------------------------------

## Installation

```{r install, eval = FALSE}
# Install from GitHub (requires devtools)
devtools::install_github("phiguera/CharAnalysis",
                         subdir = "CharAnalysis_2_0_R")

# Suggested packages for figures
install.packages(c("ggplot2", "patchwork", "ggtext"))
```

------------------------------------------------------------------------

## Worked example: Code Lake

Code Lake (site code `CO`) in Alaska, USA, is the primary validation dataset for *CharAnalysis*. The record spans approximately the past 7,300 years and is analyzed here using a local Gaussian mixture model (GMM) for threshold determination, with the working threshold percentile set at 0.95 to compensate for language-induced GMM drift relative to the MATLAB v2.0 implementation; see the *Comparison with MATLAB v2.0* section below for the rationale.

### Input files

*CharAnalysis* reads two CSV files:

- **`CO_charData.csv`** — the charcoal data table (depth, age, volume, count).
- **`CO_charParams.csv`** — the analysis parameter file.

Both files ship with the package in `inst/validation/` and are located with `system.file()`. The data file is auto-derived from the site name embedded inside the params file, so only the params path needs to be supplied to `CharAnalysis()`. In your own use you would substitute the path to a `*_charParams.csv` file of your choosing.

For this worked example we use `CO_compensated_charParams.csv`, a variant of the standard `CO_charParams.csv` in which the working threshold percentile is lowered from 0.99 to 0.95. This compensates for a language-induced shift in the GMM-fitted noise distribution; see the *Comparison with MATLAB v2.0* section below for details.

```{r paths}
params_file <- system.file("validation", "CO_compensated_charParams.csv",
                           package = "CharAnalysis")
params_file
```

### Running the full pipeline

A single call to `CharAnalysis()` runs all five analytical steps and returns a named list of results. The function prints progress messages as it works through each step.

```{r run, message = TRUE}
library(CharAnalysis)
out <- CharAnalysis(params_file)
```

### Exploring the output

`CharAnalysis()` returns a named list. The most commonly used elements are:

```{r output-structure}
names(out)
```

#### The `charcoal` object

`out$charcoal` holds all time-series outputs, both raw and processed:

```{r charcoal}
# Inspect the first few rows of key time series
head(data.frame(
  age_BP    = out$charcoal$ybpI,          # interpolated age (yr BP)
  CHAR      = out$charcoal$accI,          # C_interpolated  (pieces cm-2 yr-1)
  C_bkg     = out$charcoal$accIS,         # C_background
  C_peak    = out$charcoal$peak,          # C_peak (residuals)
  peaks     = out$charcoal$charPeaks[, 4] # final-threshold peak flags (0/1)
))
```

#### The `char_thresh` object

`out$char_thresh` holds threshold values, SNI, and goodness-of-fit results:

```{r thresh}
# Threshold at the final percentile (column 4 = threshValues[4])
range(out$char_thresh$pos[, 4], na.rm = TRUE)

# Signal-to-noise index (SNI): values > 3 indicate a strong signal
summary(out$char_thresh$SNI)
```

#### Post-processing metrics

`out$post` holds fire-history summary metrics:

```{r post}
# Fire-return intervals (FRIs) and mean FRI
cat("Number of FRIs:", length(out$post$FRI), "\n")
cat("Mean FRI:", round(mean(out$post$FRI), 1), "yr\n")

# Per-zone Weibull statistics (zone 1)
fri_z1 <- out$post$FRI_params_zone[1, ]
cat(sprintf(
  "Zone 1 — nFRI: %d  mFRI: %.1f yr  WBLb: %.1f  WBLc: %.2f\n",
  fri_z1[1], fri_z1[2], fri_z1[5], fri_z1[8]
))
```

#### The `char_results` matrix

`out$char_results` is the complete N × 33 output matrix, with columns matching the MATLAB `charResults` CSV exactly:

```{r char-results}
dim(out$char_results)

# Total number of fire events identified
sum(out$charcoal$charPeaks[, 4], na.rm = TRUE)
```

------------------------------------------------------------------------

### Recommended workflow: parameter evaluation before interpretation

CharAnalysis separates figures into diagnostic and analytical categories for a reason: the diagnostic figures are tools for deciding whether and how to interpret the analytical figures, not just documentation of what the program did. Working through them in order matters.

**Step 1: Assess interpolation and background smoothing (Figs 1 and 3)**

Inspect Fig 1 (`char_plot_raw()`) and Fig 3 (`char_plot_peaks()`) together. The central question is whether the background trend tracks genuine low-frequency variation in charcoal accumulation without fitting the peaks you are trying to detect. A smoothing window that is too narrow absorbs peaks into the background; one that is too wide misses meaningful low-frequency variation.

``` r
out <- CharAnalysis(params_file, plots = FALSE)
char_plot_raw(out)    # Fig 1: compare smoothing methods and window widths
char_plot_peaks(out)  # Fig 3: assess background fit and peak residuals
```

Adjust `smoothing$yr` and `smoothing$method` in the parameter file, re-run, and repeat until the background behavior is appropriate for the record.

**Step 2: Evaluate signal quality before proceeding (Fig 4)**

Before examining any fire-history results, inspect Fig 4 (`char_plot_sni()`). The signal-to-noise index (SNI; Kelly et al. 2011) quantifies whether the charcoal record has sufficient contrast between fire-event peaks and background noise to support statistically reliable peak detection. The recommended minimum is SNI \> 3.0.

``` r
char_plot_sni(out)   # Fig 4: SNI time series and distribution
```

**If SNI falls below 3.0 for substantial portions of the record, peak detection in those intervals is not well justified.** This is not a minor caveat. Fire events identified in low-SNI intervals cannot be distinguished from noise with confidence, and fire-history statistics derived from them — fire-return intervals, fire frequency, zone comparisons — should not be interpreted without explicit acknowledgment of this limitation. The `CharAnalysis()` console flags this condition when it occurs; take it seriously.

When SNI is \< 3, consider: (1) adjusting the smoothing window (wider windows often increase SNI by better separating signal from noise), (2) restricting interpretation to the portion of the record where SNI \> 3.0, or (3) reconsidering whether peak analysis is the appropriate method for this record.

**Step 3: Examine threshold diagnostics if needed (Fig 2)**

Fig 2 (`char_plot_thresh_diag()`) shows how the threshold is determined. For the local-threshold path (`threshType = 2`), it displays a 5×5 grid of sliding windows, each showing the C~peak~ distribution, fitted noise component, and threshold value — useful for identifying windows where the noise model fits poorly. For the global-threshold path (`threshType = 1`), it shows a single histogram of C~peak~ values with the fitted noise distribution and selected threshold. The local-path version is computationally slower and most useful when Fig 4 suggests threshold instability; it is not necessary to examine on every run.

``` r
char_plot_thresh_diag(out)   # Fig 2: threshold diagnostics
```

**Step 4: Generate analytical figures once diagnostics are satisfactory (Figs 6–9)**

Only after the background, SNI, and threshold diagnostics are acceptable should you proceed to the analytical figures.

``` r
char_plot_analysis(out)   # Figs 6-9: zone comparisons, cumulative peaks, FRI distributions, fire history
```

------------------------------------------------------------------------

### Output figures

*CharAnalysis* provides nine publication-quality ggplot2 figures organized into two categories. **Diagnostic figures (1-5)** support parameter selection and evaluation; they are produced automatically by `CharAnalysis()` (when `plots = TRUE`, the default) and can be regenerated at any time with `char_plot_diagnostic(out)`. **Analytical figures (6-9)** support fire-history interpretation and are produced by an explicit call to `char_plot_analysis(out)`. Each figure is also produced by its own `char_plot_*()` function for fine-grained control.

The figures require the **ggplot2**, **patchwork**, and **ggtext** packages, all in `Suggests`. If any of these is unavailable in the current R environment the figure chunks below will be skipped silently.

#### Diagnostic figures (Figs 1-5)

##### Figure 1 — Smoothing options for C~background~

`char_plot_raw()` overlays the raw CHAR series, the resampled (interpolated) CHAR at the constant time-step set by `yrInterp`, and the C~background~ curves produced by all five smoothing methods over the user-specified smoothing window. Inspecting these alternatives helps assess whether the chosen smoothing method is well-behaved on the record at hand and whether the smoothing window is appropriate.

```{r fig1, eval = PLOT_OK, fig.width = 7, fig.height = 5}
char_plot_raw(out)
```

##### Figure 2 — Threshold determination diagnostics

`char_plot_thresh_diag()` shows how the local threshold is determined within sliding windows along the record. The figure is organized as a 5×5 grid of windows, each panel showing the empirical C~peak~ distribution within that window, the fitted noise component (Gaussian or Gaussian mixture, per `threshMethod`), and the resulting threshold value at the working percentile.

```{r fig2, eval = PLOT_OK, fig.width = 8, fig.height = 8}
char_plot_thresh_diag(out)
```

##### Figure 3 — C~interp~, C~background~, and C~peak~

`char_plot_peaks()` shows the resampled CHAR with the fitted C~background~ trend (top panel) and the C~peak~ series with threshold lines, identified fire events, and screened candidates (bottom panel). This is typically the first figure to inspect when evaluating smoothing window and threshold settings.

```{r fig3, eval = PLOT_OK, fig.width = 7, fig.height = 6}
char_plot_peaks(out)
```

##### Figure 4 — Threshold sensitivity and SNI

`char_plot_sni()` shows sensitivity of peak identification to alternative threshold values and the signal-to-noise index (SNI) through time and as a distribution. The SNI dashed reference line at 3.0 marks the Kelly et al. (2011) minimum for reliable peak detection.

```{r fig4, eval = PLOT_OK, fig.width = 9, fig.height = 6}
char_plot_sni(out)
```

#### Analytical figures (Figs 6-9)

##### Figure 6 — Between-zone CHAR comparisons

`char_plot_zones()` tests whether raw charcoal accumulation differs between user-defined stratigraphic zones. The left panel shows empirical cumulative distribution functions of raw CHAR within each zone, with pairwise two-sample Kolmogorov-Smirnov p-values annotated. The right panel shows box plots of raw CHAR by zone.

```{r fig6, eval = PLOT_OK, fig.width = 8, fig.height = 4}
char_plot_zones(out)
```

##### Figure 7 — Cumulative peaks through time

`char_plot_cumulative()` shows the cumulative count of identified fire events plotted against age. Changes in slope identify periods of regime change in fire frequency.

```{r fig7, eval = PLOT_OK, fig.width = 7, fig.height = 4}
char_plot_cumulative(out)
```

##### Figure 8 — FRI distributions by zone

`char_plot_fri()` summarises the distribution of fire-return intervals (FRIs) within each user-defined zone. Each panel shows a histogram normalised to proportions, with a Weibull PDF overlaid where the KS goodness-of-fit test passes. Weibull scale (*b*) and shape (*c*) parameters, mean FRI, and sample size are annotated on each panel.

```{r fig8, eval = PLOT_OK, fig.width = 7, fig.height = 5}
char_plot_fri(out)
```

##### Figure 9 — Continuous fire history

`char_plot_fire_history()` is the integrated fire-history summary: peak magnitude (top), fire-return intervals through time with smoothed mean FRI and bootstrapped 95% CI (middle), and lowess-smoothed fire frequency (bottom).

```{r fig9, eval = PLOT_OK, fig.width = 7, fig.height = 8}
char_plot_fire_history(out)
```

#### Saving all figures to PDF

`char_plot_diagnostic()` and `char_plot_analysis()` each accept `save = TRUE` and `out_dir` to save their respective figure sets. `char_plot_all()` saves all nine figures in one call.

```{r save-figs, eval = FALSE}
char_plot_all(out, save = TRUE, out_dir = tempdir())
# Saves to tempdir():
#   CO_01_pretreatment.pdf           (Fig 1, when allFigures = 1)
#   CO_02_threshold_determination.pdf (Fig 2, when allFigures = 1)
#   CO_03_CHAR_analysis.pdf
#   CO_04_threshold_sensitivity_SNI.pdf
#   CO_06_zone_comparisons.pdf
#   CO_07_cumulative_peaks.pdf
#   CO_08_FRI_distributions.pdf
#   CO_09_continuous_fire_hx.pdf
```

> **Note:** The threshold-details figure (MATLAB Fig. 9) is not implemented in the R package. The background-sensitivity figure (Fig. 5, `char_bkg_sensitivity()`) is produced only when `bkgSens = 1` in the parameter file and is not shown here.

------------------------------------------------------------------------

### Writing results to CSV

`char_write_results()` writes the 33-column output matrix to a CSV file whose column names and format match the MATLAB `charResults` output exactly. `out_dir` is required (no default); substitute a path of your choosing for the temporary directory used here.

```{r write, eval = FALSE}
char_write_results(out$char_results,
                 site    = out$site,
                 out_dir = tempdir())
# Writes: <tempdir>/CO_charResults.csv
```

The output CSV contains one row per interpolated time step and 33 columns covering all analytical outputs from C~interp~ through to per-zone Weibull confidence intervals. Column headers match the MATLAB reference file exactly to facilitate direct numerical comparison.

------------------------------------------------------------------------

## Key parameters

The parameter file (`*_charParams.csv`) controls all aspects of the analysis. The most commonly adjusted parameters are:

| Parameter | Default | Description |
|----|----|----|
| `yrInterp` | 15 | Resampling resolution (yr). Set to 0 for automatic (median raw resolution). |
| `yr` | 500 | Smoothing window width (yr) for C~background~ estimation. |
| `threshType` | 2 | Threshold type: 1 = global, 2 = local (sliding window). |
| `threshMethod` | 3 | Noise distribution: 2 = Gaussian, 3 = Gaussian mixture model. |
| `threshValues` | 0.95, 0.99, 0.999, 0.99 | Percentile thresholds; the final value defines the working threshold. |
| `minCountP` | 0.05 | Alpha level for the minimum-count significance screen. |
| `peakFrequ` | 1000 | Window width (yr) for smoothed fire frequency and FRI statistics. |
| `zoneDiv` | record bounds | Zone boundaries (yr BP) for stratified FRI and Weibull analysis. |

------------------------------------------------------------------------

## Comparison with MATLAB v2.0

*CharAnalysis* in R is a direct translation of *CharAnalysis* v2.0 (MATLAB). Outputs are quantitatively equivalent on all validated reference datasets. The table below summarises results across the four validation datasets; full details are in `inst/z_Validation_report_R_vs_MATLAB_V_2.0.md`.

| Dataset | Site | Smoothing | Threshold | charBkg max\|diff\| | Peaks R v2.0.x | Peaks MATLAB v2.0 |
|----|----|----|----|----|----|----|
| CO | Code Lake, AK | Method 4 (moving median) | 0.99 | \~5 × 10^-6^ | 39 | 48 |
| CO (compensated) | Code Lake, AK | Method 4 (moving median) | 0.95\* | \~5 × 10^-6^ | 50 | 48 |
| CH10 | Chickaree Lake, CO | Method 2 (robust lowess) | 0.99 | 0.267 | 59 | 50 |
| SI17 | Silver Lake, CO | Method 2 (robust lowess) | 0.99 | 0.130 | 25 | 31 |
| RA07 | Raven Lake, AK | Method 2 (robust lowess) | 0.99 | \< 0.001 | 15 | 17 |

\* The Threshold column lists the R-side working percentile (`threshValues[4]`); MATLAB uses 0.99 in all rows. The "CO (compensated)" row therefore compares R \@ 0.95 with MATLAB \@ 0.99, illustrating the parameter offset that brings R into qualitative agreement with the published MATLAB result. See *Compensating for GMM drift* below.

Two sources of numerical difference are documented:

**1. Robust lowess background (smoothing method 2)** — MATLAB's Curve Fitting Toolbox `smooth(..., 'rlowess')` and the R `char_lowess()` port produce slightly different C~background~ series. For gap-free records (RA07) the difference is negligible (\< 0.001). For records with NaN gaps (CH10, SI17) the difference is larger (≤ 0.267) because the two implementations handle gap positions differently inside the bisquare robustness iteration. The other smoothing methods (1 plain lowess, 3 moving average, 4 moving median, 5 moving mode) are unaffected and agree to within floating-point noise on all datasets — including Code Lake (CO), which uses method 4.

**2. Gaussian mixture model (GMM) peak counts** — The R package ports the MATLAB `GaussianMixture.m` EM algorithm directly. Floating-point arithmetic accumulates differently across languages during EM iterations, causing the two implementations to reach slightly different threshold values in some local windows. Peak counts differ by 10–20% across datasets, with the direction varying (R sometimes higher, sometimes lower). All threshold and peak differences are downstream consequences of this single source; interpolation and peak-magnitude outputs agree to within numerical precision.

**Compensating for GMM drift.** The peak-count differences in the strict-comparison rows above reflect a 1-to-1 reproduction of each MATLAB v2.0 reference run, with all parameters held identical across languages. The differences are not irreducible. For Code Lake, the R implementation produces a slightly higher GMM threshold than MATLAB does, so fewer C~peak~ values pass the screen; lowering the working threshold percentile (`threshValues[4]`) from 0.99 to 0.95 compensates for this drift and brings the R peak count substantially closer to the MATLAB v2.0 count. The required compensation is dataset-specific because the direction of GMM drift varies by record. We report unmodified results in the validation table to characterize the language-induced numerical drift directly, rather than mask it.

The worked example earlier in this vignette uses the bundled `CO_compensated_charParams.csv` (identical to `CO_charParams.csv` except `threshValues[4] = 0.95`). With this compensation, R identifies 50 peaks for Code Lake. This is much closer to the published MATLAB v2.0 result of 48 peaks, and it captures the significant decrease in FRIs from Zone 1 to Zone 2, highlighted in Higuera et al. (2009).

**Weibull confidence intervals** — Bootstrap CIs use random resampling and will differ between R and MATLAB runs regardless of any other differences. Weibull point estimates (scale *b*, shape *c*) agree within a few percent on datasets where peak counts are similar.

------------------------------------------------------------------------

## Citation

If you use *CharAnalysis* in published research, please cite Higuera et al. (2009), the first study to apply the core analytical tools implemented in the program. If you used *CharAnalysis* v2.0 (MATLAB or R v2.0.x) specifically, please also cite the software:

> Higuera, P.E., L.B. Brubaker, P.M. Anderson, F.S. Hu, and T.A. Brown. 2009. Vegetation mediated the impacts of postglacial climate change on fire regimes in the south-central Brooks Range, Alaska. *Ecological Monographs* 79:201–219. <https://doi.org/10.1890/07-2019.1>

> Higuera, P.E. 2026. *CharAnalysis*: Diagnostic and analytical tools for peak analysis in sediment-charcoal records (Version 2.0). Zenodo. <https://doi.org/10.5281/zenodo.19304064>

------------------------------------------------------------------------

## References

Higuera, P.E., L.B. Brubaker, P.M. Anderson, F.S. Hu, and T.A. Brown. 2009. Vegetation mediated the impacts of postglacial climate change on fire regimes in the south-central Brooks Range, Alaska. *Ecological Monographs* 79:201–219. <https://doi.org/10.1890/07-2019.1>
